---
title: "Analysis"
author: "Ruben Props"
date: "14 januari 2017"
output:
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 2
    css: report_styles.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      cache = TRUE,
                      include = TRUE,
                      collapse = FALSE,
                      dependson = NULL,
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE,
                      fig.path="Figures/cached/",  # Set the figure options
                      fig.align = "center"
                      )
```

The full analysis of the submitted paper: *Invasive dreissenid mussels induce shifts in bacterioplankton diversity through selective feeding on high nucleic acid bacteria*. 

Before starting the analysis please unzip *16S.zip* and download the FCM data for the mussel experiment from [here](https://flowrepository.org/experiments/1034) and store them in a directory named *data_mussel*. The reference FCM data (approx. 1GB) for the cooling water system and Lake Michigan/Muskegon Lake are available from [here](https://flowrepository.org/experiments/746) and [here](https://flowrepository.org/experiments/1047). These data should be stored in a directory called *data_reference* with subdirectories *FCM_CW* for the cooling water data and *FCM_MI* for the Lake Michigan and Muskegon Lake data.

The phenotypic and taxonomic diversities (`REF_diversity.csv`, `Lakes_diversity16S_F.csv` and `Lakes_diversityFCM_F.csv`) were calculated by means of the external scripts in `/extra_scripts` as the computation can take some time.

**Notice:** The bootstrap repeats are set rather low (`R = 3`) to facilitate faster run times. For exactly reproducing the output described in the manuscript please adjust the parameters to those specified in the methods section (`R = 100`). The output from this markdown file is provided in html format.

# Accuracy
Adjust this parameter (`R` = number of bootstraps) to increase the accuracy of the phenotypic diversity estimation. For exactly reproducing the results from the manuscript set `R.main = 100`. Be aware that this will significantly increase the run time.

```{r accuracy, message=FALSE, warning = FALSE}
R.main <- 3
```
# Load libraries

```{r load-libraries, message=FALSE, warning = FALSE}
library("Phenoflow")
library("plyr")
library("dplyr")
library("gridExtra")
library("ggplot2")
library("formatR")
library("gridExtra")
library("cowplot")
library("RColorBrewer")
library("vegan")
library("sandwich")
library("lmtest")
library("grid")
library("car")
library("egg")
library("phyloseq")
library("splines")
library("caret") # for cross validation
source("functions.R")
my.settings <- list(
  strip.background=list(col="transparent"),
  strip.border=list(col="transparent", cex=5),
  gate=list(col="black", fill="lightblue", alpha=0.2,border=NA,lwd=2),
  panel.background=list(col="lightgray"),
  background=list(col="white"))
```

# Part 1: Validation of alpha diversity
```{r Regression analysis}
div.FCM <- read.csv2("files/Lakes_diversityFCM_F.csv")
div.16S <- read.csv2("files/Lakes_diversity16S_F.csv")
metadata <- read.csv2("files/Lakes_metadata.csv")
metadata <- metadata[metadata$Platform == "Accuri",]
metadata$Sample_fcm <- gsub(metadata$Sample_fcm, pattern="_rep.*", replacement="")
metadata <- do.call(rbind,by(metadata, INDICES = factor(metadata$Sample_fcm), 
                        FUN = unique))

# Calculate means + errors for FCM data
groupLevels <- factor(gsub(div.FCM$Sample_names, pattern="_rep.*", replacement="", fixed=FALSE))
means <- do.call(rbind,by(div.FCM[,c(3:5,9:12)], INDICES = groupLevels, 
                          FUN = colMeans))
errors1 <- do.call(rbind,by(div.FCM[,6:8], INDICES = groupLevels, 
                           FUN = function(x) sqrt(colSums(x^2))/ncol(x)))
errors2 <- do.call(rbind,by(div.FCM[,c(9,10,12)], INDICES = groupLevels, 
                            FUN = function(x) apply(x,2,sd)))
colnames(errors2) <- c("counts.sd", "volume.sd", "HNA_counts.sd")
div.FCM.merged <- data.frame(sample_fcm = rownames(means), cbind(means[,1:3], errors1, means[,4:7], errors2))

# Rename colnames
colnames(div.FCM.merged)[1:7] <- c("Sample_fcm","D0.fcm","D1.fcm","D2.fcm","sd.D0.fcm","sd.D1.fcm","sd.D2.fcm")

# Replace .renamed for MI13 data
div.16S$Sample <- gsub(div.16S$Sample, pattern=".renamed",replacement="",fixed=TRUE)

# Replace "-" from div.16S
div.16S$Sample <- gsub(div.16S$Sample, pattern="-", replacement="")

### Remove RNA samples
div.16S <- div.16S[sapply(as.character(div.16S$Sample), function(x) substr(x, nchar(x), nchar(x))) != "R",]

# Merge data
data.16s <- inner_join(div.16S, metadata, by=c("Sample"="Sample_16S"))

# Join all data in one dataframe 
data.16s$Sample_fcm <- gsub(data.16s$Sample_fcm, pattern="_rep.*", replacement="")
data.total <- inner_join(div.FCM.merged, data.16s, by="Sample_fcm")
  
# Select samples with > 10000 counts
data.total <- data.total[data.total$counts>10000,]

# Add fcm/16S data from previous publication (cooling water)
div.ref <- read.csv2("files/REF_diversity.csv")

data.total.final <- rbind.fill(data.total, div.ref[, c(2,4:9)])
data.total.final[(nrow(data.total)+1):nrow(data.total.final), 16:25] <- div.ref[, c(10:19)]
data.total.final$Lake <- as.character(data.total.final$Lake)
data.total.final$Lake[is.na(data.total.final$Lake)] <- "Cooling water"
data.total.final$Lake <- factor(data.total.final$Lake, levels=c("Michigan","Muskegon","Cooling water"))
data.total.final$Lake <- revalue(data.total.final$Lake, c("Michigan"="Lake Michigan", "Muskegon"="Muskegon Lake"))
lb <- read.csv2("files/REF_labels.csv")
tmp <- left_join(data.total.final,lb, by=c("Sample_fcm"="Sample_fcm"))
data.total.final$Sample[data.total.final$Lake == "Cooling water"] <- tmp$Sample_seq[!is.na(tmp$Sample_seq)]; remove(tmp)
data.total.final$Lake[1:87][data.total.final$Site[1:87] == "MLB"] <- "Muskegon Lake"
data.total.final$Season[1:87][data.total.final$Season[1:87]=="Winter"] <- "Fall"

# Chlorophyl data
Chl <- read.csv("files/Lakes_metadata_Chl.csv")
Chl <- Chl[Chl$Sample_16S %in% data.total.final$Sample[data.total.final$Lake == "Lake Michigan"],]
mean(aggregate(Chl~Sample_16S, data=Chl, mean)$Chl)
sd(aggregate(Chl~Sample_16S, data=Chl, mean)$Chl)

# Get average HNA percentage and error in lake Mi
mean(100*data.total.final$HNA_counts[data.total.final$Lake=="Lake Michigan"]/data.total.final$counts[data.total.final$Lake=="Lake Michigan"])
100*sqrt(sum((data.total.final$HNA_counts.sd[data.total.final$Lake=="Lake Michigan"]/data.total.final$HNA_counts[data.total.final$Lake=="Lake Michigan"])^2 + (data.total.final$counts.sd[data.total.final$Lake=="Lake Michigan"]/data.total.final$counts[data.total.final$Lake=="Lake Michigan"])^2)/length(data.total.final$counts[data.total.final$Lake=="Lake Michigan"]))
length(data.total.final$counts[data.total.final$Lake=="Lake Michigan"])

# Get R squared for all diversity metrics
lm.F.D2 <- lm(log2(D2)~log2(D2.fcm), data=data.total.final)
summary(lm.F.D2)$r.squared
lm.F.D1 <- lm(log2(D1)~log2(D1.fcm), data=data.total.final)
summary(lm.F.D1)$r.squared
lm.F.D0 <- lm(log2(D0)~log2(D0.fcm), data=data.total.final)
summary(lm.F.D0)$r.squared

# Calculate unbiased R squared by tenfold cross validation
lmGrid <- expand.grid(intercept = TRUE)
R.cv.D2 <- train(log2(D2)~log2(D2.fcm), data=data.total.final, method ='lm', trControl = trainControl(method ="repeatedcv", repeats = 100), tuneGrid = lmGrid)
R.cv.D2$results
R.cv.D1 <- train(log2(D1)~log2(D1.fcm), data=data.total.final, method ='lm', trControl = trainControl(method ="repeatedcv", repeats = 100), tuneGrid = lmGrid)
R.cv.D1$results
R.cv.D0 <- train(log2(D0)~log2(D0.fcm), data=data.total.final, method ='lm', trControl = trainControl(method ="repeatedcv", repeats = 100), tuneGrid = lmGrid)
R.cv.D0$results

# Dynamic range of D2
max(data.total.final$D2)/min(data.total.final$D2)
max(data.total.final$D1)/min(data.total.final$D1)

# Dynamic range of D2 in previous study
max(data.total.final$D2[data.total.final$Lake=="Cooling water"])/min(data.total.final$D2[data.total.final$Lake=="Cooling water"])
```

## Figure S1: Check model assumptions
```{r Plot D2 residuals, fig.width = 15, fig.height = 5}
# D2 and D1 are highly correlated so we only use D2 in the feeding experiment
cor(data.total.final$D1.fcm,data.total.final$D2.fcm)

# Residual analysis of the D2 model
par(mfrow=c(1,3))
qqPlot(lm.F.D2, col="blue", reps=10000, ylab="Studentized residuals", xlab="Theoretical quantiles (t-distribution)",
       cex=1.5, las=1)
plot(residuals(lm.F.D2,"pearson"), x=predict(lm.F.D2), col="blue", las=1,
     ylab="Pearson residuals",xlab="Predicted values", cex=1.5)
lines(x=c(0,10), y=c(0,0), lty=2)
plot(y=log2(data.total.final$D2), x=predict(lm.F.D2), col="blue",
     ylab="Observed values",xlab="Predicted values", cex=1.5,
     las=1)
```

## Figure 1: Regression analysis
```{r Plot D2 regression, fig.width = 8, fig.height = 5}
# Prepare to plot r squared / pearson's correlation
my_grob = grobTree(textGrob(bquote(r^2 == .(paste(round(R.cv.D2$results$Rsquared, 2)))), x=0.8,  y=0.16, hjust=0,
                            gp=gpar(col="black", fontsize=20, fontface="italic")))
my_grob2 = grobTree(textGrob(bquote(r[p] == .(round(cor(y=log2(data.total.final$D2), x=log2(data.total.final$D2.fcm)), 2))), x=0.8,  y=0.08, hjust=0,
                            gp=gpar(col="black", fontsize=20, fontface="italic")))

# Plot D2
p5 <- ggplot(data=data.total.final,aes(x=D2.fcm,y=D2, fill=Lake))+ scale_fill_manual(values=c("#88419d","#a6cee3","#fc8d62")) +
  geom_point(shape=21,size=6,alpha=0.6,aes(fill=Lake))+
  theme_bw()+labs(y=expression('Taxonomic diversity - D'[2]),x=expression('Phenotypic diversity - D'[2]), fill="Environment")+
  theme(axis.text=element_text(size=15),axis.title=element_text(size=20,face="bold"),legend.text=element_text(size=15),
        legend.title=element_text(size=16),strip.text.x = element_text(size = 22),
         legend.position = c(0.05, .98), legend.justification = c(0, 1),
        legend.background = element_rect(fill="transparent"))+
  scale_y_continuous(trans='log2', breaks = seq(5,60, 10),minor_breaks =NULL) +
  scale_x_continuous(trans='log2', breaks = seq(1000,4250,250),minor_breaks =NULL, limits = c(NA, 2700)) +
  geom_smooth(method="lm",color="black", fill ="lightblue", formula=y~x)+
  annotation_custom(my_grob)+
  annotation_custom(my_grob2)

pdf("Fig1.pdf", width = 7, height = 6)
print(p5)
dev.off()

```
## Figure S1: Compare slopes of individual regressions
```{r Compare slopes of individual regressions, fig.width = 8, fig.height = 5}
# Supplementary figure for showing that the slopes for individual regressions of each lake are not significantly different
# from each other. This allows us to make one global regression for further analysis.

p5.si <- ggplot(data=data.total.final,aes(x=D2.fcm,y=D2, fill=Lake))+ scale_fill_manual(values=c("#88419d","#a6cee3","#fc8d62")) +
  geom_point(shape=21,size=6,alpha=0.6,aes(fill=Lake))+
  theme_bw()+labs(y=expression('Taxonomic diversity - D'[2]),x=expression('Phenotypic diversity - D'[2]), fill="Environment")+
  theme(axis.text=element_text(size=15),axis.title=element_text(size=20,face="bold"),legend.text=element_text(size=15),
        legend.title=element_text(size=16),strip.text.x = element_text(size = 22))+
  scale_y_continuous(trans='log2', breaks = seq(5,60, 10),minor_breaks =NULL) +
  scale_x_continuous(trans='log2', breaks = seq(1000,4250,250),minor_breaks =NULL, limits = c(NA, 2700)) +
  geom_smooth(method="rlm",color="black",formula=y~x)

lm_comp <- rlm(log2(D2)~log2(D2.fcm)*Lake, data=data.total.final)
contrast_matrix <- rbind(
  "Cooling water vs Muskegon Lake" = c(0,0,0,0,-1,1),
  "Lake Michigan vs Cooling water" = c(0,0,0,0,0,1),
  "Lake Michigan vs Muskegon Lake" = c(0,0,0,0,1,0)
)


# Adjusted p-values using Benjamini & Hochberg (1995) correction
summary(glht(model = lm_comp, contrast_matrix), test = adjusted("BH"))

# pdf("Fig1_SI.pdf", width = 9, height = 6)
# png("Fig1_SI.png", width = 9, height = 6, res = 500, units = "in")
print(p5.si)
# dev.off()
```

## Figure S2: Regression of D0/D1
```{r Plot D0 & D1 regression, fig.width = 15, fig.height = 5}
### Prepare to plot r squared / pearson's correlation
my_grob = grobTree(textGrob(bquote(r^2 == .(round(R.cv.D1$results$Rsquared, 2))), x=0.8,  y=0.16, hjust=0,
                            gp=gpar(col="black", fontsize=20, fontface="italic")))
my_grob2 = grobTree(textGrob(bquote(r[p] == .(round(cor(y=log2(data.total.final$D1), x=log2(data.total.final$D1.fcm)), 2))), x=0.8,  y=0.08, hjust=0,
                             gp=gpar(col="black", fontsize=20, fontface="italic")))

### Plot D1
p6 <- ggplot(data=data.total.final,aes(x=D1.fcm,y=D1, fill=Lake))+ scale_fill_manual(values=c("#88419d","#a6cee3","#fc8d62")) +
  geom_point(shape=21,size=6,alpha=0.6,aes(fill=Lake))+
  theme_bw()+labs(y=expression('Taxonomic diversity - D'[1]),x=expression('Phenotypic diversity - D'[1]), fill="Environment")+
  theme(axis.text=element_text(size=15),axis.title=element_text(size=18,face="bold"),legend.text=element_text(size=15), legend.title=element_text(size=16),strip.text.x = element_text(size = 22))+
  scale_y_continuous(trans='log2', breaks = seq(20,200, 40),minor_breaks =NULL) +
  scale_x_continuous(trans='log2', breaks = seq(1500,4250,500),minor_breaks =NULL) +
  geom_smooth(method="lm",color="black",fill="lightblue",formula=y~x)+
  annotation_custom(my_grob)+
  annotation_custom(my_grob2)

### Prepare to plot r squared / pearson's correlation
my_grob = grobTree(textGrob(bquote(r^2 == .(round(R.cv.D0$results$Rsquared, 2))), x=0.8,  y=0.16, hjust=0,
                            gp=gpar(col="black", fontsize=20, fontface="italic")))
my_grob2 = grobTree(textGrob(bquote(r[p] == .(round(cor(y=log2(data.total.final$D0), x=log2(data.total.final$D0.fcm)), 2))), x=0.8,  y=0.08, hjust=0,
                             gp=gpar(col="black", fontsize=20, fontface="italic")))

### Plot D0
p7 <- ggplot(data=data.total.final,aes(x=D0.fcm,y=D0,fill=Lake))+ scale_fill_manual(values=c("#88419d","#a6cee3","#fc8d62")) +
  geom_point(shape=21,size=6,alpha=0.6,aes(fill=Lake))+
  theme_bw()+labs(y=expression('Taxonomic diversity - D'[0]),x=expression('Phenotypic diversity - D'[0]),
                  fill="Environment")+
  theme(axis.text=element_text(size=15),axis.title=element_text(size=18,face="bold"),legend.text=element_text(size=15),
        legend.title=element_text(size=16),strip.text.x = element_text(size = 22))+
  scale_y_continuous(trans='log2', breaks = seq(0,4000, 500),minor_breaks =NULL) +
  scale_x_continuous(trans='log2', breaks = seq(1500,20000,2500),minor_breaks =NULL) +
  geom_smooth(method="lm",color="black",fill="lightblue",formula=y~x)+
  annotation_custom(my_grob)+
  annotation_custom(my_grob2)

### All together
grid.arrange(p7, p6, ncol=2)
```

## Figure S4: Effect of sample size
```{r sample-size, fig.width = 15, fig.height = 5, message = FALSE, warning = FALSE}
# Load data
path = "data_mussel"
flowData_sc <- read.flowSet(path = path, transformation = FALSE, pattern=".fcs")

# Preprocess data according to standard protocol
flowData_sc <- transform(flowData_sc, `FL1-H` = asinh(`FL1-H`), `SSC-H` = asinh(`SSC-H`), 
                                  `FL3-H` = asinh(`FL3-H`), `FSC-H` = asinh(`FSC-H`))
param = c("FL1-H", "FL3-H", "SSC-H", "FSC-H")
flowData_sc = flowData_sc[, param]

# Test this for only one sample
flowData_sc <- flowData_sc[10]

# Create a PolygonGate for denoising the dataset Define coordinates for
# gate in sqrcut1 in format: c(x,x,x,x,y,y,y,y)
sqrcut1 <- matrix(c(8.5,8.5,15,15,3,8,14,3),ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H", "FL3-H")
polyGate1 <- polygonGate(.gate = sqrcut1, filterId = "Total Cells")

# Isolate only the cellular information based on the polyGate1
flowData_sc<- Subset(flowData_sc, polyGate1)

summary <- fsApply(x = flowData_sc, FUN = function(x) apply(x, 2, max), use.exprs = TRUE)
max = max(summary[, 1])
mytrans <- function(x) x/max
flowData_sc <- transform(flowData_sc, `FL1-H` = mytrans(`FL1-H`), 
                                  `FL3-H` = mytrans(`FL3-H`), `SSC-H` = mytrans(`SSC-H`), `FSC-H` = mytrans(`FSC-H`))

# Subsample at various depths and calculate diversity metrics with 100
# bootstraps Notice: this will use some CPU/RAM
for (i in c(10, 100, 200, 300, 400, 500, 750, 1000, 1250, 1500, 2000, 2500, 
            3000, 5000, 10000, 15000, 20000, 30000)) {
  for (j in 1:100) {
    fs1 <- FCS_resample(flowData_sc, replace = TRUE, sample = i, progress = FALSE)
    fp <- flowBasis(fs1, param, nbin = 128, bw = 0.01, normalize = function(x) x)
    div.tmp <- Diversity(fp, d = 3, R = 100, progress = FALSE)
    div.tmp <- cbind(div.tmp, size = i)
    if (j == 1) 
      results <- div.tmp else results <- rbind(results, div.tmp)
  }
  if (i == 10) 
    results.tot <- results else results.tot <- rbind(results.tot, results)
}

# Create plots
D0 <- ggplot(data = results.tot, aes(x = factor(size), y = D0)) + # geom_jitter(alpha=0.7, size=1)+
  geom_boxplot(alpha = 0.2, color = "blue", fill = "blue", size = 1) + 
  labs(x = "Sample size (nr. of cells)", y = expression('Phenotypic diversity - D'[0])) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45,hjust = 1),
                     axis.text=element_text(size=11),axis.title=element_text(size=15))

D1 <- ggplot(data = results.tot, aes(x = factor(size), y = D1)) + # geom_jitter(alpha=0.7, size=1)+
  geom_boxplot(alpha = 0.2, color = "blue", fill = "blue", size = 1) + 
  labs(x = "Sample size (nr. of cells)", y = expression('Phenotypic diversity - D'[1])) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.text=element_text(size=11), axis.title=element_text(size=15))

D2 <- ggplot(data = results.tot, aes(x = factor(size), y = D2)) + # geom_jitter(alpha=0.7, size=1)+
  geom_boxplot(alpha = 0.2, color = "blue", fill = "blue", size = 1) + 
  labs(x = "Sample size (nr. of cells)", y = expression('Phenotypic diversity - D'[2])) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.text=element_text(size=11), axis.title=element_text(size=15))

# Sample used for this assessment
grid.arrange(D0, D1, D2, ncol = 3, top = textGrob(paste("Sample used:",flowCore::sampleNames(flowData_sc)),                                                   gp = gpar(fontsize = 20, font = 3)))
```

# Part 2: Validation of beta diversity
```{r beta-diversity analysis}
myColours2 <- brewer.pal(n=12,"Paired"); myColours2 <- myColours2[c(2,6,7)]

# Import otu data
physeq.otu <- import_mothur(mothur_shared_file ="16S/stability.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.an.unique_list.shared" ,
                            mothur_constaxonomy_file = "16S/stability.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.an.unique_list.0.03.cons.taxonomy")
otu_table(physeq.otu) <- t(otu_table(physeq.otu))
physeq.otu <- prune_samples(sample_sums(physeq.otu)>10000, physeq.otu)
physeq.otu <- prune_taxa(taxa_sums(physeq.otu)>0, physeq.otu)

# Select samples for which you have FCM data
sample_names(physeq.otu) <- gsub(sample_names(physeq.otu), pattern=".renamed", replacement="")
sample_names(physeq.otu) <- gsub(sample_names(physeq.otu), pattern=".renamed", replacement="")
meta.seq <- read.csv2("files/Lakes_metadata.csv")
physeq.otu <- prune_samples(sample_names(physeq.otu) %in% data.total.final$Sample, physeq.otu)

# Annotate with metadata
rownames(data.total.final) <- data.total.final$Sample
sample_data(physeq.otu) <- data.total.final

# Rescale
sample_sums(physeq.otu)
physeq.otu <- scale_reads(physeq.otu)
sample_sums(physeq.otu)
physeq.otu <- transform_sample_counts(physeq.otu, function(x) x/sum(x))

# Run beta diversity analysis on 16s data
pcoa <- ordinate(
  physeq = physeq.otu, 
  method = "PCoA", 
  distance = "bray",
  correction = "lingoes",
  k=2
)

pcoa.df <- data.frame(pcoa$vectors, sample_data(physeq.otu))
var <- round(pcoa$values$Eigenvalues/sum(pcoa$values$Eigenvalues)*100,1)

# Start beta diversity analysis on FCM data
path = "data_reference/FCM_MI"
flowData_transformed  <- read.flowSet(path = path, transformation = FALSE, pattern=".fcs")

flowData_transformed <- transform(flowData_transformed,`FL1-H`=asinh(`FL1-H`), 
                                  `SSC-H`=asinh(`SSC-H`), 
                                  `FL3-H`=asinh(`FL3-H`), 
                                  `FSC-H`=asinh(`FSC-H`))
param=c("FL1-H", "FL3-H","SSC-H","FSC-H")
flowData_transformed = flowData_transformed[,param]

# Create a PolygonGate for denoising the dataset
# Define coordinates for gate in sqrcut1 in format: c(x,x,x,x,y,y,y,y)
sqrcut1 <- matrix(c(8.5,8.5,15,15,3,8,14,3),ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
polyGate1 <- polygonGate(.gate=sqrcut1, filterId = "Total Cells")

#  Gating quality check
xyplot(`FL3-H` ~ `FL1-H`, data=flowData_transformed[1], filter=polyGate1,
       scales=list(y=list(limits=c(0,14)),
                   x=list(limits=c(6,16))),
       par.settings = my.settings,
       axis = axis.default, nbin=125, 
       par.strip.text=list(col="white", font=2, cex=2), smooth=FALSE)

# Isolate only the cellular information based on the polyGate1
flowData_transformed <- Subset(flowData_transformed, polyGate1)

summary <- fsApply(x=flowData_transformed,FUN=function(x) apply(x,2,max),use.exprs=TRUE)
max = max(summary[,1])
mytrans <- function(x) x/max
flowData_transformed <- transform(flowData_transformed,`FL1-H`=mytrans(`FL1-H`),
                                  `FL3-H`=mytrans(`FL3-H`), 
                                  `SSC-H`=mytrans(`SSC-H`),
                                  `FSC-H`=mytrans(`FSC-H`))

# Calculate fingerprint with bw = 0.01
fbasis <- flowBasis(flowData_transformed, param, nbin=128, 
                    bw=0.01,normalize=function(x) x)

# Run beta diversity
pos <- gsub(rownames(fbasis@basis),pattern="_rep.*", replacement="")
beta.div <- beta_div_fcm(fbasis, INDICES=pos, ord.type="PCoA")
df.beta.fcm <- data.frame(Sample = rownames(beta.div$points), beta.div$points)
df.beta.fcm$Sample <- gsub(df.beta.fcm$Sample, pattern="MI5", replacement = "M15")
df.beta.fcm <- inner_join(df.beta.fcm, data.total.final, by=c("Sample"="Sample_fcm"))
var2 <- round(beta.div$eig/sum(beta.div$eig)*100, 1)
df.beta.fcm <- droplevels(df.beta.fcm)

# Procrustes analysis
fbasis1 <- fbasis
fbasis1@basis <- fbasis1@basis/apply(fbasis1@basis, 1, max)
fbasis1@basis <- round(fbasis1@basis, 3)
x <- by(fbasis1@basis, INDICES = pos, FUN = colMeans)
x <- do.call(rbind, x)
rownames(x) <- gsub(rownames(x), pattern = "MI5", replacement = "M15")
x <- x[(rownames(x) %in% data.total.final$Sample_fcm), ]

# Rename and order rows of fcm/seq data
tmp <- data.frame(sample_fcm=rownames(x))
tmp <- left_join(tmp, data.total.final, by=c("sample_fcm"="Sample_fcm"))
rownames(x) <- tmp$Sample; remove(tmp)
x <- x[order(rownames(x)),]
x.data <- data.total.final[data.total.final$Sample %in% rownames(x),]
x.data <- x.data[order(x.data$Sample),]
otu_table(physeq.otu) <- otu_table(physeq.otu)[order(rownames(otu_table(physeq.otu))),]

# Run PcoA
dist.fcm <- vegan::vegdist(x)
pcoa.fcm <- cmdscale(dist.fcm)
dist.seq <- vegan::vegdist(otu_table(physeq.otu))
pcoa.seq <- cmdscale(dist.seq)

```

## Procrustes analysis
```{r procrustes-analysis, fig.width = 5, fig.height = 5}
# Run procrustes + permutation
# Permutations are constrained within each sampling year 
perm <- how(nperm = 999)
setBlocks(perm) <- with(x.data, Year)
dist.prot <- vegan::protest(pcoa.seq,pcoa.fcm, permutations = perm)
summary(dist.prot)
plot(dist.prot)
```

## Figure 2: Beta diversity analysis
```{r permanova-analysis, fig.width = 12, fig.height = 5}
# Run PERMANOVA to evaluate if conclusions are similar between FCM/seq
# Similar variances across seasons
disper.fcm <- betadisper(dist.fcm, group=x.data$Season)
print(disper.fcm)
disper.seq <- betadisper(dist.seq, group=x.data$Season)
print(disper.seq)

# Permutations are constrained within each sampling year 
perm <- how(nperm = 999)
setBlocks(perm) <- with(x.data, Year)
permanova.fcm <- adonis(dist.fcm ~ Season * Lake, data = x.data, 
    permutations = perm)
permanova.seq <- adonis(dist.seq ~ Season * Lake, data = data.frame(sample_data(physeq.otu)), 
    permutations = perm)

# Plot FCM beta diversity
my_grob = grobTree(textGrob(bquote(paste(r[Season]^2 == 
    .(round(100 * permanova.fcm$aov.tab[1, 5], 1)), 
    "%")), x = 0.7, y = 0.95, hjust = 0, gp = gpar(col = "black", 
    fontsize = 14, fontface = "italic")))
my_grob2 = grobTree(textGrob(bquote(paste(r[Lake]^2 == 
    .(format(round(100 * permanova.fcm$aov.tab[2, 5], 
        1), nsmall = 1)), "%")), x = 0.7, y = 0.87, 
    hjust = 0, gp = gpar(col = "black", fontsize = 14, 
        fontface = "italic")))
my_grob3 = grobTree(textGrob(bquote(paste(r[Season:Lake]^2 == 
    .(round(100 * permanova.fcm$aov.tab[3, 5], 1)), 
    "%")), x = 0.7, y = 0.79, hjust = 0, gp = gpar(col = "black", 
    fontsize = 14, fontface = "italic")))

df.beta.fcm$Season <- factor(df.beta.fcm$Season, levels = c("Spring", "Summer", "Fall"))
pcoa.df$Season <- factor(pcoa.df$Season, levels = c("Spring", "Summer", "Fall"))

beta.pcoa.fcm <- ggplot(data=df.beta.fcm, aes(x=X1, y=-X2, shape=Lake))+
  scale_shape_manual(values=c(21,24))+
  geom_point(size=7, aes(fill = Season), alpha=0.7)+
  theme_bw()+
  scale_fill_manual(values=myColours2)+
  scale_colour_manual(values=myColours2)+
  labs(x = paste0("PCoA axis 1 (",var2[1], "%)"), y = paste0("PCoA axis 2 (",var2[2], "%)"), fill="", shape = "", colour="",
       title="B")+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=20),
        title=element_text(size=20), legend.text=element_text(size=16))+ 
  guides(fill = guide_legend(override.aes = list(shape = 22)))+
  annotation_custom(my_grob)+
  annotation_custom(my_grob2)+
  annotation_custom(my_grob3)

# Plot 16S beta diversity
my_grob = grobTree(textGrob(bquote(paste(r[Season]^2 == 
    .(round(100 * permanova.seq$aov.tab[1, 5], 1)), 
    "%")), x = 0.7, y = 0.95, hjust = 0, gp = gpar(col = "black", 
    fontsize = 14, fontface = "italic")))
my_grob2 = grobTree(textGrob(bquote(paste(r[Lake]^2 == 
    .(round(100 * permanova.seq$aov.tab[2, 5], 1)), 
    "%")), x = 0.7, y = 0.87, hjust = 0, gp = gpar(col = "black", 
    fontsize = 14, fontface = "italic")))
my_grob3 = grobTree(textGrob(bquote(paste(r[Season:Lake]^2 == 
    .(round(100 * permanova.seq$aov.tab[3, 5], 1)), 
    "%")), x = 0.7, y = 0.79, hjust = 0, gp = gpar(col = "black", 
    fontsize = 14, fontface = "italic")))

beta.pcoa <- ggplot(data=pcoa.df, aes(x=Axis.1, y=Axis.2, shape=Lake))+
  geom_point(alpha=0.7, size=7, aes(fill=Season))+
  scale_shape_manual(values=c(21,24))+
  theme_bw()+
  scale_fill_manual(values=myColours2)+
  scale_colour_manual(values=myColours2)+
  labs(x = paste0("PCoA axis 1 (",var[1], "%)"), y = paste0("PCoA axis 2 (",var[2], "%)"), fill="", shape="", colour="",
       title="A")+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=20),
        title=element_text(size=20), legend.text=element_text(size=16))+
  guides(fill = guide_legend(override.aes = list(shape = 22)))+
  annotation_custom(my_grob)+
  annotation_custom(my_grob2)+
  annotation_custom(my_grob3)+
  ylim(-0.4,0.3)

# Both beta diversity plots together
grid_arrange_shared_legend(beta.pcoa, beta.pcoa.fcm, ncol=2)

```

# Part 3: Mussel experiment

```{r Mussel-fcm-analysis}
# Set seed for reproducible analysis
set.seed(777)

myColours <- brewer.pal("Accent",n=3)

# Samples were diluted 2x
dilution <- 2

path = "data_mussel"
flowData <- read.flowSet(path = path, transformation = FALSE, pattern=".fcs")

flowData_transformed <- transform(flowData,`FL1-H`=asinh(`FL1-H`), 
                                  `SSC-H`=asinh(`SSC-H`), 
                                  `FL3-H`=asinh(`FL3-H`), 
                                  `FSC-H`=asinh(`FSC-H`))
param=c("FL1-H", "FL3-H","SSC-H","FSC-H")
flowData_transformed = flowData_transformed[,param]
remove(flowData)

# Create a PolygonGate for denoising the dataset
# Define coordinates for gate in sqrcut1 in format: c(x,x,x,x,y,y,y,y)
sqrcut1 <- matrix(c(8.5,8.5,15,15,3,8,14,3),ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
polyGate1 <- polygonGate(.gate=sqrcut1, filterId = "Total Cells")

#  Gating quality check
xyplot(`FL3-H` ~ `FL1-H`, data=flowData_transformed[1], filter=polyGate1,
       scales=list(y=list(limits=c(0,14)),
                   x=list(limits=c(6,16))),
       par.settings = my.settings,
       axis = axis.default, nbin=125, 
       par.strip.text=list(col="white", font=2, cex=2), smooth=FALSE)

# Isolate only the cellular information based on the polyGate1
flowData_transformed <- Subset(flowData_transformed, polyGate1)

summary <- fsApply(x=flowData_transformed,FUN=function(x) apply(x,2,max),use.exprs=TRUE)
max = max(summary[,1])
mytrans <- function(x) x/max
flowData_transformed <- transform(flowData_transformed,`FL1-H`=mytrans(`FL1-H`),
                                  `FL3-H`=mytrans(`FL3-H`), 
                                  `SSC-H`=mytrans(`SSC-H`),
                                  `FSC-H`=mytrans(`FSC-H`))


# Calculate fingerprint with bw = 0.01
fbasis <- flowBasis(flowData_transformed, param, nbin=128, 
                    bw=0.01,normalize=function(x) x)

# Calculate ecological parameters from normalized fingerprint 
# Densities will be normalized to the interval [0,1]
# d = rounding factor
# Diversity.fbasis <- Diversity(fbasis, d = 3, plot = TRUE, R = 999)
Diversity.fbasis <- Diversity_rf(flowData_transformed, d=3, param = param, R = R.main)


# make metadata table
tmp <- strsplit(rownames(Diversity.fbasis)[1:42], "_")
meta.div <- cbind(do.call(rbind, lapply(tmp, rbind))[,1],rep("C",42), do.call(rbind, lapply(tmp, rbind))[,2:3])
tmp <- strsplit(rownames(Diversity.fbasis)[43:nrow(Diversity.fbasis)], "_")
meta.div <- data.frame(rbind(meta.div, do.call(rbind, lapply(tmp, rbind))))
colnames(meta.div) <- c("Sample","Treatment","Time","Replicate")
meta.div$Replicate <- gsub(meta.div$Replicate,pattern=".fcs",replacement="")
meta.div$Time <- as.numeric(gsub(meta.div$Time,pattern="t",replacement=""))


# Merge with Diversity.fbasis
Diversity.fbasis <- cbind(Diversity.fbasis, meta.div)

# Remove outliers due to human errors (forgot staining)
fbasis@basis <- fbasis@basis[Diversity.fbasis$D2>1600,]
flowData_transformed <- flowData_transformed[Diversity.fbasis$D2>1600]

sqrcut1 <- matrix(c(asinh(12500),asinh(12500),15,15,3,9.55,14,3)/max,ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
rGate_HNA <- polygonGate(.gate=sqrcut1, filterId = "HNA")
sqrcut1 <- matrix(c(8.5,8.5,asinh(12500),asinh(12500),3,8,9.55,3)/max,ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
rGate_LNA <- polygonGate(.gate=sqrcut1, filterId = "LNA")

# Diversities of HNA/LNA populations
flowData_HNA <- split(flowData_transformed, rGate_HNA)$`HNA+`
flowData_LNA <- split(flowData_transformed, rGate_LNA)$`LNA+`

Diversity.HNA <- Diversity_rf(flowData_HNA, d=3, param = param, R = R.main)
Diversity.LNA <- Diversity_rf(flowData_LNA, d=3, param = param, R = R.main)

# Counts
# Normalize total cell gate
sqrcut1 <- matrix(c(8.5,8.5,15,15,3,8,14,3)/max,ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
polyGate1 <- polygonGate(.gate=sqrcut1, filterId = "Total Cells")

# Check if rectangle gate is correct, if not, adjust rGate_HNA
xyplot(`FL3-H` ~ `FL1-H`, data=flowData_transformed[1], filter=rGate_HNA,
       scales=list(y=list(limits=c(0,1)),
                   x=list(limits=c(0.4,1))),
       par.settings = my.settings,
       axis = axis.default, nbin=125, par.strip.text=list(col="white", font=2, 
                                                          cex=2), smooth=FALSE)
# Extract the cell counts
a <- flowCore::filter(flowData_transformed, rGate_HNA) 
HNACount <- flowCore::summary(a);HNACount <- toTable(HNACount)
s <- flowCore::filter(flowData_transformed, polyGate1)
TotalCount <- flowCore::summary(s);TotalCount <- flowCore::toTable(TotalCount)

# Extract the volume
vol <- c()
for(i in 1:length(flowData_transformed)){
  vol[i] <- as.numeric(flowData_transformed[[i]]@description$`$VOL`)/1000
}

# Save counts
counts <- data.frame(Samples=flowCore::sampleNames(flowData_transformed), 
                     Total.cells = dilution*TotalCount$true/vol, HNA.cells = dilution*HNACount$true/vol,
                     LNA.cells = dilution*(TotalCount$true-HNACount$true)/vol)

# Pool results
Diversity.fbasis <- Diversity.fbasis[Diversity.fbasis$D2>1600,]
Storage <- c(); Storage[Diversity.fbasis$Treatment=="T"] <- "Covered"; Storage[Diversity.fbasis$Treatment=="S"] <- "Submerged"
results <- cbind(Diversity.fbasis,counts, Storage, Diversity.HNA, Diversity.LNA)

# Add an extra column for biological replicates
bio_rep <- c(rep(1,21),rep(2,21),rep(1,41),rep(2,41),rep(3,41))
results <- cbind(results,bio_rep=factor(bio_rep))

# Select only S experiment for beta-div and further analysis

levels(results$Treatment)[levels(results$Treatment)=="C"] <- "Control"
levels(results$Treatment)[levels(results$Treatment)=="T"] <- "Feeding - T"
levels(results$Treatment)[levels(results$Treatment)=="S"] <- "Feeding - S"

# Beta diversity analysis
pos <- gsub(rownames(fbasis@basis),pattern="_rep1", replacement="")
pos <- gsub(pos,pattern="_rep2", replacement="")
pos <- gsub(pos,pattern="_rep3", replacement="")
pos <- factor(pos)
tmp <- results[,c(8,9,10,16,31)]
tmp <- by(tmp, INDICES=pos, FUN=unique)
tmp <- do.call(rbind, tmp)

# All samples
beta.div <- beta_div_fcm(fbasis, INDICES=pos, ord.type="PCoA")
fbasis1<-fbasis; fbasis1@basis <- fbasis@basis[!Diversity.fbasis$Treatment=="T",]
fbasis2<-fbasis; fbasis2@basis <- fbasis@basis[!Diversity.fbasis$Treatment=="S",]

# Without T and S versus control separately
beta.div.S <- beta_div_fcm(fbasis1, INDICES=pos[!Diversity.fbasis$Treatment=="T"], ord.type="PCoA")
# Take average fingerprint for technical replicates
fbasis1@basis <- fbasis1@basis/apply(fbasis1@basis, 1, max)
fbasis1@basis <- round(fbasis1@basis, 4)
x <- by(fbasis1@basis, INDICES = pos[!Diversity.fbasis$Treatment=="T"], FUN = colMeans)
x <- do.call(rbind, x)
# Calculate distance matrix
dist.S <- vegdist(x, method="bray")

beta.div.T <- beta_div_fcm(fbasis2, INDICES=pos[!Diversity.fbasis$Treatment=="S"], ord.type="PCoA")


results <- data.frame(cbind(Sample_names=levels(pos), do.call(rbind,by(results[,c(2,3,4,13,14,15,18,19,20,25,26,27)], INDICES=pos, FUN=colMeans)), 
                   do.call(rbind,by(results[,c(5,6,7,21,22,23,28,29,30)], INDICES=pos, FUN = function(x) sqrt(colSums(x^2))/nrow(x))),
                   do.call(rbind,by(results[,c(13,14,15)], INDICES=pos, FUN = function(x) apply(x,2,sd))),
                   tmp))

colnames(results)[colnames(results)=="Total.cells.1"|colnames(results)=="HNA.cells.1"|colnames(results)=="LNA.cells.1"] <- 
  c("sd.Total.cells","sd.HNA.cells","sd.LNA.cells")

colnames(results)[colnames(results)=="sd.D0.1"|colnames(results)=="sd.D1.1"|colnames(results)=="sd.D2.1"] <- 
  c("sd.D0.HNA","sd.D1.HNA","sd.D2.HNA")
colnames(results)[colnames(results)=="D0.1"|colnames(results)=="D1.1"|colnames(results)=="D2.1"] <- 
  c("D0.HNA","D1.HNA","D2.HNA")

colnames(results)[colnames(results)=="sd.D0.2"|colnames(results)=="sd.D1.2"|colnames(results)=="sd.D2.2"] <- 
  c("sd.D0.LNA","sd.D1.LNA","sd.D2.LNA")
colnames(results)[colnames(results)=="D0.2"|colnames(results)=="D1.2"|colnames(results)=="D2.2"] <- 
  c("D0.LNA","D1.LNA","D2.LNA")

# Import metadata of the mussels for M&M
meta.mus <- read.csv2("files/Experiment_metadata_mussels.csv")
meta.mus <- meta.mus[meta.mus$Treatment=="S",]
mean(meta.mus$size_mm)
sd(meta.mus$size_mm)

# No significant difference between mesocosms in mussel size distribution (p=0.081).
kruskal.test(size_mm ~ Sample, data=meta.mus)

# Subset results for mussels transported by the best method (submerged)
result.tmp <- results
results <- result.tmp[!result.tmp$Treatment=="Feeding - T", ]
results$Treatment <- droplevels(plyr::revalue(results$Treatment, c("Feeding - S"="Feeding")))
```

## Diversity dynamics during IDM feeding

```{r fit-splines}
# Calculate the dynamics of D2 in time for treatment/control
# We can't use a linear regression here (for obvious reasons)
# Hence we will try to fit robust smoothings splines and see if we can make inference this way
sp_T <- rlm(D2~splines::ns(Time, df=3), data=results[results$Treatment=="Feeding",])
sp_C <- rlm(D2~splines::ns(Time, df=3), data=results[results$Treatment=="Control",])

# Order studentized residuals according to timepoint
res.T <- residuals(sp_T, "pearson")[order(results[results$Treatment=="Feeding",]$Time)]
res.C <- residuals(sp_C, "pearson")[order(results[results$Treatment=="Control",]$Time)]

# Location of knots
attr(splines::ns(results$Time, df=3), "knots")
```

## Figure S5: Check for autocorrelation
```{r evaluate auto-correlation, fig.width = 12, fig.height = 5}
# Check for temporal autocorrelation in model residuals
par(mfrow=c(1,2))

acf(res.C, main="Treatment: control", las=1)
acf(res.T, main="Treatment: feeding", las=1)

# Perform statistical inference on splines
car::Anova(sp_C, vcov = vcovHAC(sp_C)) # p = 0.03795
car::Anova(sp_T, vcov = vcovHAC(sp_T)) # p = 5.599e-06

# Compare errors before and after correction
  # Treatment
summary(sp_T)$coefficients[,2]
sqrt(diag(vcovHAC(sp_T)))
  # Control
summary(sp_C)$coefficients[,2]
sqrt(diag(vcovHAC(sp_C)))
``` 

## Figure 3: Diversity analysis during feeding
```{r plot-diversity, fig.width = 12, fig.height = 5}
# Prepare plots
p.alpha <- ggplot(data=results, aes(x=Time, y=D2, fill=Treatment)) + 
  geom_point(shape=21, size=7,alpha=0.9)+
  scale_fill_manual(values=myColours[c(1,2)])+
  theme_bw()+
  labs(y=expression('Phenotypic diversity - D'[2]), x="Time (h)", title="A",
       fill="")+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=20),
        title=element_text(size=20), legend.text=element_text(size=16),
        legend.direction = "horizontal",legend.position = "bottom")+ 
  geom_smooth(method="rlm", color="black", alpha=0.2, formula = y ~ splines::ns(x,df=3))+
  ylim(1990,2150)

# Separate for submerged data
# PERMANOVA on beta diversity analysis
disper.test <- betadisper(dist.S, group=results$Treatment)
disper.test # average distance to mean 0.03 for both groups
anova(disper.test) # P = 0.892
perma.beta <- adonis(dist.S~Treatment*Time, data=results)
my_grob = grobTree(textGrob(bquote(paste(r[Feeding]^2 == .(round(100 * perma.beta$aov.tab[1, 
    5], 1)), "%")), x = 0.68, y = 0.95, hjust = 0, gp = gpar(col = "black", fontsize = 14, 
    fontface = "italic")))
my_grob2 = grobTree(textGrob(bquote(paste(r[Time]^2 == .(format(round(100 * perma.beta$aov.tab[2, 
    5], 1), nsmall = 1)), "%")), x = 0.68, y = 0.87, hjust = 0, gp = gpar(col = "black", 
    fontsize = 14, fontface = "italic")))
my_grob3 = grobTree(textGrob(bquote(paste(r[Feeding:Time]^2 == .(round(100 * perma.beta$aov.tab[3, 
    5], 1)), "%")), x = 0.68, y = 0.79, hjust = 0, gp = gpar(col = "black", fontsize = 14, 
    fontface = "italic")))

beta.div.data.S <- data.frame(beta.div.S$points, tmp[!tmp$Treatment=="Feeding - T",])
beta.div.data.S <- droplevels(beta.div.data.S)
beta.div.data.S$Treatment <- plyr::revalue(beta.div.data.S$Treatment, c("Feeding - S"="Feeding"))
var <- round(vegan::eigenvals(beta.div.S)/sum(vegan::eigenvals(beta.div.S))*100,1)

p.beta.S <- ggplot(data=beta.div.data.S, aes(x=X1, y=X2, fill=Treatment, size=Time))+
  geom_point(shape=21, alpha=1)+
  scale_size(range=c(4,10), breaks=c(0,0.5,1,1.5,2,2.5,3))+ 
  guides(fill = FALSE, size = guide_legend(nrow = 2))+
  theme_bw()+
  scale_fill_manual(values=myColours[c(1,2)])+
  labs(x = paste0("PCoA axis 1 (",var[1], "%)"), y = paste0("PCoA axis 2 (",var[2], "%)"), title="B",
       fill="", size = "Time (h)")+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=20),
        title=element_text(size=20), legend.text=element_text(size=16),
        legend.direction = "horizontal",
        legend.position = c(0.15, -0.13), 
        legend.justification = c(0, 1),
        legend.background = element_rect(fill="transparent"))+
  annotation_custom(my_grob)+
  annotation_custom(my_grob2)+
  annotation_custom(my_grob3)+
  ylim(-0.035,0.045)

### Plot diversity dynamics
p.beta.S.plot <-  p.beta.S + theme(legend.position = "none")
p.alpha.plot <-  p.alpha + theme(legend.position = "none")

pdf("Fig3_revised.pdf", width = 12, height = 6)
grid.arrange(p.alpha.plot, p.beta.S, g_legend(p.alpha), layout_matrix = rbind(c(1,2),
                                                                               c(3,3)),
             heights=c(6, 1))
dev.off()
```

## Infer taxonomic diversity dynamics
```{r calculate-diversity-dynamics}
# D2 regression from part I
lm.F <- lm(log2(D2)~log2(D2.fcm), data=data.total.final)

# Calculate mean predicted decrease in D2 at 0 and 3 hours
df <- data.frame(D2.fcm = results$D2[results$Treatment=="Feeding"][results[results$Treatment=="Feeding",]$Time==3 | results[results$Treatment=="Feeding",]$Time==0],
                 time=results$Time[results$Treatment=="Feeding"][results[results$Treatment=="Feeding",]$Time==3 | results[results$Treatment=="Feeding",]$Time==0])
df.pred <- data.frame(D2.16S = predict(lm.F, df, se=TRUE)$fit, D2.16S.error = predict(lm.F, df, se=TRUE)$se.fit, time = df$time)
df.pred <- data.frame(aggregate(D2.16S~time, df.pred, mean), aggregate(D2.16S.error~time, df.pred, FUN = function(x) sqrt(sum(x^2))/length(x)))

# Note: After transformation back to linear scale, errors are not necessarily normal distributed anymore, so calculate minimum and maximum and report as such.
df.pred.res <- df.pred 
# Maximum decrease in D2 (3.63 units or 15.6 %)
2^(df.pred$D2.16S[df.pred.res$time==0] + df.pred$D2.16S.error[df.pred.res$time==0]) -  2^(df.pred$D2.16S[df.pred.res$time==3] - df.pred$D2.16S.error[df.pred.res$time==3])
(2^(df.pred$D2.16S[df.pred.res$time==0] + df.pred$D2.16S.error[df.pred.res$time==3]) -  2^(df.pred$D2.16S[df.pred.res$time==3] - df.pred$D2.16S.error[df.pred.res$time==3])) / 2^(df.pred$D2.16S[df.pred.res$time==0] + df.pred$D2.16S.error[df.pred.res$time==0])

# Minimum decrease in D2 (1.65 units or 7.5 %)
2^(df.pred$D2.16S[df.pred.res$time==0] - df.pred$D2.16S.error[df.pred.res$time==0]) -  2^(df.pred$D2.16S[df.pred.res$time==3] + df.pred$D2.16S.error[df.pred.res$time==3])
(2^(df.pred$D2.16S[df.pred.res$time==0] - df.pred$D2.16S.error[df.pred.res$time==3]) -  2^(df.pred$D2.16S[df.pred.res$time==3] + df.pred$D2.16S.error[df.pred.res$time==3])) / 2^(df.pred$D2.16S[df.pred.res$time==0] - df.pred$D2.16S.error[df.pred.res$time==0])

# Average decrease in D2 (2.64 units or 11.6 %)
2^df.pred$D2.16S[df.pred.res$time==0] -  2^df.pred$D2.16S[df.pred.res$time==3]
(2^df.pred$D2.16S[df.pred.res$time==0] -  2^df.pred$D2.16S[df.pred.res$time==3]) / 2^df.pred$D2.16S[df.pred.res$time==0] 

# Note: Errors seem (approx.) randomly distributed even on linear scale, so we can just summarize by +/- the average on the mean estimate.

# Calculate mean predicted decrease in D2 at 0 and 1 hour
df <- data.frame(D2.fcm = results$D2[results$Treatment=="Feeding"][results[results$Treatment=="Feeding",]$Time==1 | results[results$Treatment=="Feeding",]$Time==0],
                 time=results$Time[results$Treatment=="Feeding"][results[results$Treatment=="Feeding",]$Time==1 | results[results$Treatment=="Feeding",]$Time==0])
df.pred <- data.frame(D2.16S = predict(lm.F, df, se=TRUE)$fit, D2.16S.error = predict(lm.F, df, se=TRUE)$se.fit, time = df$time)
df.pred <- data.frame(aggregate(D2.16S~time, df.pred, mean), aggregate(D2.16S.error~time, df.pred, FUN = function(x) sqrt(sum(x^2))/length(x)))

# Note: After transformation back to linear scale, errors are not necessarily normal distributed anymore, so calculate minimum and maximum and report as such.

df.pred.res <- df.pred 
# Maximum decrease in D2 (2.82 units or 12.05 %)
2^(df.pred$D2.16S[df.pred.res$time==0] + df.pred$D2.16S.error[df.pred.res$time==0]) -  2^(df.pred$D2.16S[df.pred.res$time==1] - df.pred$D2.16S.error[df.pred.res$time==1])
(2^(df.pred$D2.16S[df.pred.res$time==0] + df.pred$D2.16S.error[df.pred.res$time==1]) -  2^(df.pred$D2.16S[df.pred.res$time==1] - df.pred$D2.16S.error[df.pred.res$time==1])) / 2^(df.pred$D2.16S[df.pred.res$time==0] + df.pred$D2.16S.error[df.pred.res$time==0])

# Minimum decrease in D2 (0.79 units or 3.66 %)
2^(df.pred$D2.16S[df.pred.res$time==0] - df.pred$D2.16S.error[df.pred.res$time==0]) -  2^(df.pred$D2.16S[df.pred.res$time==1] + df.pred$D2.16S.error[df.pred.res$time==1])
(2^(df.pred$D2.16S[df.pred.res$time==0] - df.pred$D2.16S.error[df.pred.res$time==1]) -  2^(df.pred$D2.16S[df.pred.res$time==1] + df.pred$D2.16S.error[df.pred.res$time==1])) / 2^(df.pred$D2.16S[df.pred.res$time==0] - df.pred$D2.16S.error[df.pred.res$time==0])

# Average decrease in D2 (1.8 units or 7.96 %)
2^df.pred$D2.16S[df.pred.res$time==0] -  2^df.pred$D2.16S[df.pred.res$time==1]
(2^df.pred$D2.16S[df.pred.res$time==0] -  2^df.pred$D2.16S[df.pred.res$time==1]) / 2^df.pred$D2.16S[df.pred.res$time==0] 
# Calculate delta in D2.fcm for an increase of 10 taxonomic units (5->10 and 35->45)
df2 <- data.frame(D2.fcm=c(1000,1500,2000,2500))
df2.res <- predict(lm.F,df2)
# at low diversities
2^(df2.res[2] - df2.res[1])
# at higher diversities
2^(df2.res[4] - df2.res[3])
```

## Figure S3: Gating strategy
```{r Gating, fig.width = 15, fig.height = 10}
path = "data_mussel"
flowData <- read.flowSet(path = path, transformation = FALSE, pattern=".fcs")

flowData_transformed <- transform(flowData,`FL1-H`=asinh(`FL1-H`), 
                                  `SSC-H`=asinh(`SSC-H`), 
                                  `FL3-H`=asinh(`FL3-H`), 
                                  `FSC-H`=asinh(`FSC-H`))
param=c("FL1-H", "FL3-H","SSC-H","FSC-H")
flowData_transformed = flowData_transformed[,param]
remove(flowData)

# Create a PolygonGate for denoising the dataset
# Define coordinates for gate in sqrcut1 in format: c(x,x,x,x,y,y,y,y)
sqrcut1 <- matrix(c(8.5,8.5,15,15,3,8,14,3),ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
polyGate1 <- polygonGate(.gate=sqrcut1, filterId = "Total Cells")

summary <- fsApply(x=flowData_transformed,FUN=function(x) apply(x,2,max),use.exprs=TRUE)
max = max(summary[,1])
max <- 14.99341
mytrans <- function(x) x/max
flowData_transformed <- transform(flowData_transformed,`FL1-H`=mytrans(`FL1-H`),
                                  `FL3-H`=mytrans(`FL3-H`), 
                                  `SSC-H`=mytrans(`SSC-H`),
                                  `FSC-H`=mytrans(`FSC-H`))

flowData_transformed <- flowData_transformed[which(flowCore::sampleNames(flowData_transformed) 
                                                   %in% c("Q3_T_t0_rep1.fcs","Q3_T_t1.5_rep1.fcs","Q3_T_t3_rep1.fcs",
                                                         "C1_t0_rep1.fcs","C1_t1.5_rep1.fcs","C1_t3_rep1.fcs"))]

MyText <- c("Control - 0h", "Control - 1.5h", "Control - 3h",
            "Feeding - 0h", "Feeding - 1.5h", "Feeding - 3h")

sqrcut1 <- matrix(c(asinh(12500),asinh(12500),15,15,3,9.55,14,3)/max,ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
rGate_HNA <- polygonGate(.gate=sqrcut1, filterId = "HNA")
sqrcut1 <- matrix(c(8.5,8.5,asinh(12500),asinh(12500),3,8,9.55,3)/max,ncol=2, nrow=4)
colnames(sqrcut1) <- c("FL1-H","FL3-H")
rGate_LNA <- polygonGate(.gate=sqrcut1, filterId = "LNA")

filters <- filters(list(rGate_LNA, rGate_HNA))
flist <- list(filters , filters, filters, filters, filters, filters)
names(flist) <- flowCore::sampleNames(flowData_transformed)

print(xyplot(`FL3-H`~`FL1-H`, data=flowData_transformed,index.cond=list(c(1:6)),
             filter=flist,
             xbins=200,nbin=128, par.strip.text=list(col="black", font=3,cex=1.85), 
             smooth=FALSE, xlim=c(0.5,1),ylim=c(0.1,1),xlab=list(label="Green fluorescence intensity (FL1-H)",cex=2),ylab=list(label="Red fluorescence intensity (FL3-H)",cex=2),
             par.settings=my.settings,
             scales=list(x=list(at=seq(from=0, to=1, by=.1),cex=1),
                         y=list(at=seq(from=0, to=1, by=.2),cex=1)), layout=c(3,2),
             strip=strip.custom(factor.levels=MyText),
             margin=TRUE,
             binTrans="log"
      )
)
```

## Figure 4: Identify physiological populations
``` {r Contrasts, fig.width = 15, fig.height = 12, warning = FALSE}
comp_t0 <- fp_contrasts(x=fbasis, comp2=(pos=="C1_t0.fcs" | pos== "C2_t0.fcs"),
                        comp1=(pos=="Q1_T_t0.fcs" | pos== "Q2_T_t0.fcs" | pos== "Q3_T_t0.fcs"), 
                        thresh=0.04)
comp_t1.5 <- fp_contrasts(x=fbasis, comp2=(pos=="C1_t1.5.fcs" | pos== "C2_t1.5.fcs"),
                          comp1=(pos=="Q1_T_t1.5.fcs" | pos== "Q2_T_t1.5.fcs" | pos== "Q3_T_t1.5.fcs"), 
                          thresh=0.04)
comp_t3 <- fp_contrasts(x=fbasis, comp2=(pos=="C1_t3.fcs" | pos== "C2_t3.fcs"),
                        comp1=(pos=="Q1_T_t3.fcs" | pos== "Q2_T_t3.fcs" | pos== "Q3_T_t3.fcs"), 
                        thresh=0.04)

# Merge data frames
comp_total <- rbind(comp_t0, comp_t1.5, comp_t3)
comp_total <- data.frame(comp_total, Timepoint=as.factor(c(rep("0h Feeding", nrow(comp_t0)), 
                                                 rep("1.5h Feeding", nrow(comp_t1.5)),
                                                 rep("3h Feeding", nrow(comp_t3)))))

p12b <- ggplot(data=results, aes(x=Time, y=HNA.cells, fill=Treatment)) + 
  # geom_boxplot(alpha=0.9)+
  geom_point(shape=21, size=5,alpha=0.9)+
  scale_fill_manual(values=myColours[c(1,2)])+
  theme_bw()+
  labs(y=expression("HNA cells µL"^"-1"), x="Time (h)", title="C", fill="")+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=16),
        title=element_text(size=20), legend.text=element_text(size=14),
        legend.title=element_text(size=15),
        # panel.grid.major = element_blank(), panel.grid.minor = element_blank()
        legend.direction = "horizontal",legend.position = "bottom",
        panel.grid.minor = element_blank()
  )+ 
  # guides(fill=FALSE)+
  geom_errorbar(aes(ymin=HNA.cells-sd.HNA.cells, ymax=HNA.cells+sd.HNA.cells), width=0.075)+
  ylim(200,525)+ 
  geom_smooth(method="rlm",color="black", alpha=0.2)+
  scale_x_continuous(breaks=c(0,0.5,1,1.5,2,2.5,3))

p13 <- ggplot(data=results, aes(x=Time, y=LNA.cells, fill=Treatment)) + 
  # geom_boxplot(alpha=0.9)+
  geom_point(shape=21, size=5,alpha=0.9)+
  scale_fill_manual(values=myColours[c(1,2)])+
  # geom_smooth(formula=y ~ x, color="black")+
  # geom_boxplot(mapping=factor(Time),alpha=0.4,outlier.shape=NA)+
  theme_bw()+
  labs(y=expression("LNA cells µL"^"-1"), x="Time (h)", title="B", fill="")+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=16),
        title=element_text(size=20), legend.text=element_text(size=14),
        legend.title=element_text(size=15),
        legend.direction = "horizontal",legend.position = "bottom",
        panel.grid.minor = element_blank())+ 
  # guides(fill=FALSE)+
  geom_errorbar(aes(ymin=LNA.cells-sd.LNA.cells, ymax=LNA.cells+sd.LNA.cells), width=0.075)+
  ylim(1000,1325)+
  scale_x_continuous(breaks=c(0,0.5,1,1.5,2,2.5,3))

# Contrast plot
vtot <- ggplot(comp_total, aes(`FL1.H`, `FL3.H`, z = Density))+
  geom_tile(aes(fill=Density)) + 
  geom_point(colour="gray", alpha=0.4)+
  scale_fill_distiller(palette="RdBu", na.value="white") + 
  stat_contour(aes(fill=..level..), geom="polygon", binwidth=0.1)+
  theme_bw()+
  geom_contour(color = "white", alpha = 1)+
  facet_grid(~Timepoint)+
  labs(x="Green fluorescence intensity (a.u.)", y="Red fluorescence intensity (a.u.)",title="A")+
  theme(axis.title=element_text(size=16), strip.text.x=element_text(size=16),
        legend.title=element_text(size=15),legend.text=element_text(size=14),
        axis.text = element_text(size=14),title=element_text(size=20),
        strip.background=element_rect(fill=adjustcolor("lightgray",0.2))
        #,panel.grid.major = element_blank(), panel.grid.minor = element_blank()
        )

# Make the plot
g1 <- ggplotGrob(p13);
g2 <- ggplotGrob(p12b);
gv <- ggplotGrob(vtot);
fg1 <- gtable_frame(g1)
fg2 <- gtable_frame(g2)
fg12 <- gtable_frame(cbind(fg1,fg2))
fg3 <- gtable_frame(gv)
grid.newpage()
combined <- rbind(fg3, fg12)
grid.draw(combined)
```

## Calculate removal statistics
```{r HNA removal statistics}
# Calculate coefficient of variation within LNA and DNA of control/treatment
# CV for control in LNA population
100*sd(results$LNA[results$Treatment=="Control"])/mean(results$LNA[results$Treatment=="Control"])
# CV for treatment in LNA population
100*sd(results$LNA[results$Treatment=="Feeding"])/mean(results$LNA[results$Treatment=="Feeding"])
# CV for control in HNA population
100*sd(results$HNA[results$Treatment=="Control"])/mean(results$HNA[results$Treatment=="Control"])
# CV for treatment in HNA population
100*sd(results$HNA[results$Treatment=="Feeding"])/mean(results$HNA[results$Treatment=="Feeding"])

# Calculate the removal rate of HNA bacteria
# We use robust regression because of the two (suspected) outliers at t0.
lm.HNA <- rlm(HNA.cells~Time, data=results[results$Treatment=="Feeding",])
lm.HNA_C <- rlm(HNA.cells~Time, data=results[results$Treatment=="Control",])
car::Anova(lm.HNA_C) # p = 0.9826
car::Anova(lm.HNA) # p = 9.02e-11
# To get std. error on parameter estimation
summary(lm.HNA)

# Average HNA removal relative to HNA pool
t0 <- mean(results$HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="0"])
t3 <- mean(results$HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="3"])
100*(t0-t3)/t0

# Error on average HNA densities
e0 <- sqrt(sum(results$sd.HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="0"]^2))/3
e3 <- sqrt(sum(results$sd.HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="3"]^2))/3

# Error on HNA removal ratio
100*sqrt((sqrt(e0^2 + e3^2)/(t0-t3))^2 + (e0/t0)^2)

# Average HNA removal relative to total cell pool
t0 <- mean(results$HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="0"])
t3 <- mean(results$HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="3"])
t.C <- mean(results$Total.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="3"])
100*(t0-t3)/t.C

# Error on average HNA densities
e0 <- sqrt(sum(results$sd.HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="0"]^2))/3
e3 <- sqrt(sum(results$sd.HNA.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="3"]^2))/3
e.C <- sqrt(sum(results$sd.Total.cells[results$Treatment=="Feeding"][results$Time[results$Treatment=="Feeding"]=="3"]^2))/3

# Error on HNA removal ratio
100*sqrt((sqrt(e0^2 + e3^2)/(t0-t3))^2 + (e.C/t.C)^2)

# Rate per mg DW of IDM and associated error
1000*lm.HNA$coefficients[2]/mean(1000*meta.mus$weight_g)
sqrt((sd(meta.mus$weight_g)/mean(meta.mus$weight_g))^2 + (summary(lm.HNA)$coefficients[2,2]/lm.HNA$coefficients[2])^2)

# Calculage clearance rate (linear regression used to get dC/dt)
# Volume = 11L for each bucket
11*abs(lm.HNA$coefficients[2])/lm.HNA$coefficients[1]/mean(meta.mus$weight_g)
# error on first quotient
eq1 <- sqrt((summary(lm.HNA)$coefficients[1,2]/lm.HNA$coefficients[1])^2 + (summary(lm.HNA)$coefficients[2,2]/lm.HNA$coefficients[2])^2)
# error on final clearance rate
11*sqrt(((eq1/abs(lm.HNA$coefficients[2])/lm.HNA$coefficients[1])^2) + (sd(meta.mus$weight_g)/mean(meta.mus$weight_g))^2)
```
